---
output: github_document
theme: united
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```
# Statistical Rethinking

<!-- badges: start -->
<!-- badges: end -->

My notes, code, examples and problem sets for the textbook [Statistical Rethinking - *A Bayesian Course with Examples in R and Stan* by Richard McElreath](https://xcelab.net/rm/statistical-rethinking/)"  The .md documents for each chapter are readable.  

### Course: 
[updated 2022 lectures](https://www.youtube.com/playlist?list=PLDcUM9US4XdMROZ57-OIRtIK0aOynbgZN)   

[**Chapter 2**](https://github.com/MattPM/Rthink/blob/master/Chapter%202.%20Small%20Worlds%20and%20Large%20Worlds/ch2.md)  
[**Chapter 3**](https://github.com/MattPM/Rthink/blob/master/Chapter%203.%20Sampling%20the%20Imaginary/ch3.md)  
[**Chapter 4**](https://github.com/MattPM/Rthink/blob/master/Chapter%204.%20Geocentric%20Models/Ch4.md)  
[**Chapter 5**](https://github.com/MattPM/Rthink/blob/master/Chapter%205.%20DAGs%20multiple%20regression%20causal%20inference/ch5.md)  
[**Chapter 6**](https://github.com/MattPM/Rthink/blob/master/Chapter%206.%20bias%20colineariry%20berkson's%2C%20simpsons%20paradox%20types%20of%20bias/ch6.md)  
[**Chapter 7**](https://github.com/MattPM/Rthink/blob/master/Chapter%207.%20model%20selection%20information%20theory%20entropy%20cross%20validation/ch7.md)  
[**Chapter 8**](https://github.com/MattPM/Rthink/blob/master/chapter%208.%20interactions%20discrete%20and%20continuous/ch8.md)  
[**Chapter 9**](https://github.com/MattPM/Rthink/blob/master/Chapter%209.%20Markov%20Chain%20Monte%20Carlo/ch9.md)  
[**Chapter 10**](https://github.com/MattPM/Rthink/blob/master/Chapter%2010.%20GLMs%20and%20maximum%20entropy/ch10.md)  
[**Ch 11 part I: GLMs and binomial regression**](https://github.com/MattPM/Rthink/blob/master/Chapter%2011.%20count%20mods%20link%20funs%20Binom%20logistic%20Pois%20NB%20Multinom/ch11.1_GLM_binomial.md)  
[**Ch 11 part II: Poisson and Gamma Poisson (NB) regression**](https://github.com/MattPM/Rthink/blob/master/Chapter%2011.%20count%20mods%20link%20funs%20Binom%20logistic%20Pois%20NB%20Multinom/ch11.2_poisson.md)  
[**Ch 11 part III: Multinomial and categorical models**](https://github.com/MattPM/Rthink/blob/master/Chapter%2011.%20count%20mods%20link%20funs%20Binom%20logistic%20Pois%20NB%20Multinom/ch11.3_multinomial.md)  
[**Ch 11 p set**](https://github.com/MattPM/Rthink/blob/master/Chapter%2011.%20count%20mods%20link%20funs%20Binom%20logistic%20Pois%20NB%20Multinom/ch11_pset.md)  
<!-- [**Chapter 12**]()   -->
<!-- [**Chapter 13**]()   -->
<!-- [**Chapter 14**]()   -->
<!-- [**Chapter 15**]()   -->

```{r}
suppressMessages(library(rethinking))
set.seed(1990)
```

## Quick notes for using rethinking package tools

### sim some data  
```{r, fig.width=3, fig.height=3}
N = 50
z = rnorm(n = N, mean = 0, sd = 0.2)
x = z + rnorm(n = N, mean = 0, sd = 0.1)
y = x + rnorm(n = N, mean = 2,sd = 0.3)
d = data.frame(x,z,y)
pairs(d, col = rangi2)
```

### create dag 

```{r, fig.width=2, fig.height=2}
suppressMessages(library(dagitty))
dag = dagitty::dagitty(
"dag{
x -> y <- z
}")
drawdag(dag)
impliedConditionalIndependencies(dag)
adjustmentSets(dag, exposure = 'x', outcome = 'y')
```


### Specify model formula and fit using mcmc (or quadratic approximation)
```{r}
f1 = alist( 
  # estimand
  y ~ dnorm(mu, sigma),
  # likelihood
  mu <- alpha + Bx*x, 
  # priors 
  alpha ~ dnorm(0,0.5), 
  Bx ~ dnorm(0,0.5), 
  sigma ~ dexp(1)
  )

# fit model 
m1 = quap(flist = f1,data = d)
# m1 = ulam(flist = f1,data = d)

# summary of model fit
precis(m1, depth = 2)
```

### calculate and visualize uncertainty in parameter estimates and prediction intervals of simulated data based on model

```{r, fig.width=4, fig.height=4}
# specify interval of predictor values to simulate over
# multivariate models -- include other predictors 
xseq = seq(-2,2, length.out = 30)

# summarize estimand based on new predictor values
mu = link(m1, data = data.frame(x = xseq))
mu.mean = apply(mu, 2, mean)
mu.PI = apply(mu, 2, PI)

# simulate uncertainty in estimates based on full model 
dsim = sim(fit = m1, data = data.frame(x = xseq))
sim.pi = apply(dsim, 2, PI)

# plot summary
plot(y ~ x, d, col = rangi2, pch = 16)
lines(x = xseq,y = mu.mean, lwd = 3) # mean
shade(object = mu.PI, lim = xseq) # mean uncertainty
shade(object = sim.pi,lim = xseq)
```


## Other topics

### prior predictive simulation 
```{r, fig.width=3, fig.height=3}

# m1 = quap(flist, data)
# simulate prior from model formula 
p = extract.prior(m1)

# range of values for standrdized area to simulate priors over 
xseq = c(-2,2)

# simulate posterior distribution of mu
# data arg must be named as all predictors in the model, here just `x`
post.sim = link(fit = m1, post = p, data = list(x = c(-2,2)))

# plot implied prosterior lines 
plot(NULL, xlim = xseq, ylim = xseq)
for (i in 1:100) {
  lines(xseq, post.sim[i, ], col = col.alpha('black', alpha = 0.1))
}

```

### inspect posterior estimates
```{r}
# note set depth = 2 for more complex models 
precis(m1)
```

### View stan code for the model 

```{r, eval = FALSE}
stancode(m1)
```

the output of `stancode(m1)` is copied below: 

```{stan, eval=FALSE, output.var = 'test'}
data{
    int<lower=1> N;
    real y[N];
    real x[N];
}
parameters{
    real alpha;
    real Bx;
    real<lower=0> sigma;
}
model{
    vector[N] mu;
    sigma ~ exponential( 1 );
    Bx ~ normal( 0 , 0.5 );
    alpha ~ normal( 0 , 0.5 );
    for ( i in 1:N ) {
        mu[i] = alpha + (Bx * x[i]);
    }
    y ~ normal( mu , sigma );
}
generated quantities{
    vector[N] mu;
    for ( i in 1:N ) {
        mu[i] = alpha + (Bx * x[i]);
    }
}
```


### Lectures  
Lecture playlist on Youtube: <[Statistical Rethinking 2022](https://www.youtube.com/playlist?list=PLDcUM9US4XdMROZ57-OIRtIK0aOynbgZN)>

[//]: # (11 Feb SPP conflict , 25 Feb Winter Break conflict )

| Week ## | Meeting date | Reading | Lectures |
| ------- | -------------- | ------------- | ---------------------- |
| Week 01 | 07 January  | Chapters 1, 2 and 3 | [1] <[The Golem of Prague](https://youtu.be/cclUd_HoRlo)> <[(Slides)](https://speakerdeck.com/rmcelreath/statistical-rethinking-2022-lecture-01)> <br> [2] <[Bayesian Inference](https://www.youtube.com/watch?v=guTdrfycW2Q&list=PLDcUM9US4XdMROZ57-OIRtIK0aOynbgZN&index=2)> <[(Slides)](https://speakerdeck.com/rmcelreath/statistical-rethinking-2022-lecture-02)> 
| Week 02 | 14 January | Chapters 4 and 5 | [3] <[Basic Regression](https://www.youtube.com/watch?v=zYYBtxHWE0A)> <[(Slides)](https://speakerdeck.com/rmcelreath/statistical-rethinking-2022-lecture-03)> <br> [4] <[Categories & Curves](https://youtu.be/QiHKdvAbYII)> <[(Slides)](https://speakerdeck.com/rmcelreath/statistical-rethinking-2022-lecture-04)>
| Week 03 | 21 January | Chapters 5 and 6 |  [5] <[Elemental Confounds](https://youtu.be/UpP-_mBvECI)> <[(Slides)](https://speakerdeck.com/rmcelreath/statistical-rethinking-2022-lecture-05)> <br> [6] <[Good & Bad Controls](https://www.youtube.com/watch?v=NSuTaeW6Orc&list=PLDcUM9US4XdMROZ57-OIRtIK0aOynbgZN&index=6)> <[(Slides)](https://speakerdeck.com/rmcelreath/statistical-rethinking-2022-lecture-06)>
| Week 04 | 28 January | Chapters 7, 8 and 9 | [7] <[Overfitting](https://www.youtube.com/watch?v=odGAAJDlgp8&list=PLDcUM9US4XdMROZ57-OIRtIK0aOynbgZN&index=7)> <[(Slides)](https://speakerdeck.com/rmcelreath/statistical-rethinking-2022-lecture-07)> <br> [8] <[Markov chain Monte Carlo](https://www.youtube.com/watch?v=Qqz5AJjyugM&list=PLDcUM9US4XdMROZ57-OIRtIK0aOynbgZN&index=8&pp=sAQB)> <[(Slides)](https://speakerdeck.com/rmcelreath/statistical-rethinking-2022-lecture-08)>
| Week 05 | 04 February | Chapters 10 and 11 | [9] <[Logistic and Binomial GLMs](https://www.youtube.com/watch?v=nPi5yGbfxuo&list=PLDcUM9US4XdMROZ57-OIRtIK0aOynbgZN&index=9)> <[(Slides)](https://speakerdeck.com/rmcelreath/statistical-rethinking-2022-lecture-09)> <br> [10] <[Sensitivity and Poisson GLMs](https://www.youtube.com/watch?v=YrwL6t0kW2I&list=PLDcUM9US4XdMROZ57-OIRtIK0aOynbgZN&index=10)> <[(Slides)](https://speakerdeck.com/rmcelreath/statistical-rethinking-2022-lecture-10)>
| Week 06 | 11 February | Chapters 12 and 13 | [11] <[Ordered Categories](https://www.youtube.com/watch?v=-397DMPooR8&list=PLDcUM9US4XdMROZ57-OIRtIK0aOynbgZN&index=11)> <[(Slides)](https://speakerdeck.com/rmcelreath/statistical-rethinking-2022-lecture-11)> <br> [12] <[Multilevel Models](https://www.youtube.com/watch?v=SocRgsf202M&list=PLDcUM9US4XdMROZ57-OIRtIK0aOynbgZN&index=12)> <[(Slides)](https://speakerdeck.com/rmcelreath/statistical-rethinking-2022-lecture-12)>
| Week 07 | 18 February | Chapters 13 and 14 | [13] <[Multi-Multilevel Models](https://www.youtube.com/watch?v=n2aJYtuGu54&list=PLDcUM9US4XdMROZ57-OIRtIK0aOynbgZN&index=13)> <[(Slides)](https://speakerdeck.com/rmcelreath/statistical-rethinking-2022-lecture-13)> <br> [14] <[Correlated varying effects](https://www.youtube.com/watch?v=XDoAglqd7ss&list=PLDcUM9US4XdMROZ57-OIRtIK0aOynbgZN&index=14)> <[(Slides)](https://speakerdeck.com/rmcelreath/statistical-rethinking-2022-lecture-14)>
| Week 08 | 25 February | Chapter 14 | [15] <[Social Networks](https://www.youtube.com/watch?v=L_QumFUv7C8&list=PLDcUM9US4XdMROZ57-OIRtIK0aOynbgZN&index=15)> <[(Slides)](https://speakerdeck.com/rmcelreath/statistical-rethinking-2022-lecture-15)> <br> [16] <[Gaussian Processes](https://www.youtube.com/watch?v=PIuqxOBJqLU&list=PLDcUM9US4XdMROZ57-OIRtIK0aOynbgZN&index=16)> <[(Slides)](https://speakerdeck.com/rmcelreath/statistical-rethinking-2022-lecture-16)>
| Week 09 | 04 March | Chapter 15 | [17] <[Measurement Error](https://www.youtube.com/watch?v=lTFAB6QmwHM&list=PLDcUM9US4XdMROZ57-OIRtIK0aOynbgZN&index=17)> <[(Slides)](https://speakerdeck.com/rmcelreath/statistical-rethinking-2022-lecture-17)> <br> [18] <[Missing Data](https://www.youtube.com/watch?v=oMiSb8GKR0o&list=PLDcUM9US4XdMROZ57-OIRtIK0aOynbgZN&index=18)> <[(Slides)](https://speakerdeck.com/rmcelreath/statistical-rethinking-2022-lecture-18)>
| Week 10 | 11 March | Chapters 16 and 17 | [19] <[Beyond GLMs](https://www.youtube.com/watch?v=Doaod09YitA&list=PLDcUM9US4XdMROZ57-OIRtIK0aOynbgZN&index=19)> <[(Slides)](https://speakerdeck.com/rmcelreath/statistical-rethinking-2022-lecture-19)> <br> [20] <[Horoscopes](https://www.youtube.com/watch?v=qV3eos7maGs&list=PLDcUM9US4XdMROZ57-OIRtIK0aOynbgZN&index=20)> <[(Slides)](https://speakerdeck.com/rmcelreath/statistical-rethinking-2022-lecture-20)>

